---
title: "Advanced features"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced features}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r setup}
library(ggplot2)
library(marimekko)

titanic <- as.data.frame(Titanic)
```

This vignette covers the advanced features of `marimekko` beyond the
basics shown in `vignette("getting-started")`.

## Basic marimekko plot

```{r basic}
ggplot(titanic) +
  geom_marimekko(aes(fill = Survived, weight = Freq), formula = ~ Class | Survived) +
  scale_x_marimekko() +
  labs(
    title = "Marimekko plot",
    y = "Proportion"
  )
```

## Pearson residuals

Pearson residuals measure how much each cell deviates from the
independence assumption. Positive residuals indicate more observations
than expected; negative residuals indicate fewer.

Residuals are automatically computed and exposed as the
`.resid` computed variable, which you can map to an aesthetic via
`after_stat()`:

```{r residuals}
ggplot(titanic) +
  geom_marimekko(
    aes(
      fill = Survived, weight = Freq,
      alpha = after_stat(abs(.resid))
    ),
    formula = ~ Class | Survived
  ) +
  scale_x_marimekko() +
  scale_alpha_continuous(range = c(0.3, 1), guide = "none") +
  labs(title = "Residual shading: stronger opacity = larger deviation")
```

You can also map residuals to colour instead of relying on fill:

```{r residuals-colour}
ggplot(titanic) +
  geom_marimekko(aes(fill = Survived, weight = Freq),
    formula = ~ Class | Survived
  ) +
  geom_marimekko_text(aes(
    x = Class, fill = Survived, weight = Freq,
    label = after_stat(round(.resid, 1))
  ), residuals = TRUE, colour = "white", size = 3) +
  scale_x_marimekko() +
  labs(title = "Pearson residuals as labels")
```

## Jittered points

For small-to-medium datasets, `geom_marimekko_jitter()` scatters
individual data points inside their corresponding mosaic tile. Each
row is expanded by its weight, so weighted data shows the correct
number of points.

```{r jitter}
ucb <- as.data.frame(UCBAdmissions)
ucb_a <- ucb[ucb$Dept == "A", ]

ggplot(ucb_a) +
  geom_marimekko(aes(fill = Admit, weight = Freq),
    formula = ~ Gender | Admit, alpha = 0.3
  ) +
  geom_marimekko_jitter(aes(x = Gender, fill = Admit, weight = Freq),
    seed = 42, size = 0.8
  ) +
  scale_x_marimekko() +
  labs(title = "UCB admissions, Dept A: individual applicants")
```

Set `seed` for reproducible jitter placement.

## Three-variable nested mosaic

`geom_marimekko()` supports multi-variable formulas. A three-variable
formula (`~ X | Y | Z`) partitions the plot in alternating directions
(horizontal, vertical, horizontal):

- First split: horizontal by `X` (column widths proportional to `X`)
- Second split: vertical by `Y` within each column
- Third split: horizontal by `Z` within each cell

```{r multi}
ggplot(titanic) +
  geom_marimekko(aes(fill = Survived, weight = Freq),
    formula = ~ Class | Survived | Sex
  ) +
  scale_x_marimekko() +
  labs(title = "Nested mosaic: Class > Sex > Survived")
```

This produces a richer view than faceting because all three variables
share a single coordinate space, making relative proportions directly
comparable.

## Y-axis scale

By default the y-axis shows proportions from 0 to 1.
`scale_y_marimekko()` replaces that with fill category labels at
segment midpoints:

```{r y-scale}
ggplot(titanic) +
  geom_marimekko(aes(fill = Survived, weight = Freq),
    formula = ~ Class | Survived
  ) +
  scale_x_marimekko() +
  scale_y_marimekko() +
  labs(title = "Fill labels on y-axis")
```

This is most useful when you want to de-emphasize the numeric
proportions and instead label segments directly.

## Data extraction with fortify

`fortify_marimekko()` returns computed tile positions as a plain data
frame without creating a plot. This is useful for:

- Custom downstream analysis
- Manual ggplot2 layer construction
- Exporting tile coordinates

```{r fortify}
tiles <- fortify_marimekko(titanic, Class, Survived, weight = Freq)
head(tiles)
```

With residuals:

```{r fortify-resid}
tiles_resid <- fortify_marimekko(titanic, Class, Survived,
  weight = Freq, residuals = TRUE
)
head(tiles_resid)
```

The returned columns are:

| Column | Description |
|--------|-------------|
| `x_label` | Category name for the x variable |
| `fill_label` | Category name for the fill variable |
| `xmin`, `xmax` | Horizontal extent of the tile |
| `ymin`, `ymax` | Vertical extent of the tile |
| `x`, `y` | Tile center coordinates |
| `weight` | Aggregated count |
| `cond_prop` | Conditional proportion within the column |
| `.resid` | Pearson residual (when `residuals = TRUE`) |

You can use the fortified data frame to build completely custom plots:

```{r fortify-custom}
tiles <- fortify_marimekko(titanic, Class, Survived, weight = Freq)

ggplot(tiles) +
  geom_rect(
    aes(
      xmin = xmin, xmax = xmax,
      ymin = ymin, ymax = ymax,
      fill = fill_label
    ),
    colour = "grey30", linewidth = 0.3
  ) +
  geom_text(aes(x = x, y = y, label = weight), size = 3) +
  theme_minimal() +
  labs(fill = "Survived")
```

## Combining layers

Because `marimekko` produces standard ggplot2 layers, you can freely
combine multiple features:

```{r combined}
ggplot(titanic) +
  geom_marimekko(
    aes(
      fill = Survived, weight = Freq,
      alpha = after_stat(abs(.resid))
    ),
    formula = ~ Class | Survived
  ) +
  geom_marimekko_text(aes(
    x = Class, fill = Survived, weight = Freq,
    label = after_stat(weight)
  ), colour = "white", size = 3.5) +
  scale_x_marimekko(show_percentages = TRUE) +
  scale_alpha_continuous(range = c(0.4, 1), guide = "none") +
  theme_marimekko() +
  labs(
    title = "Full-featured mosaic plot",
    subtitle = "Residual shading + counts + marginal %",
    y = "Proportion"
  )
```

## Independent x/y gaps

By default, `gap` controls both horizontal (between columns) and
vertical (between segments) spacing. Use `gap_x` and `gap_y` to set
them independently:

```{r gap-xy}
ggplot(titanic) +
  geom_marimekko(aes(fill = Survived, weight = Freq),
    formula = ~ Class | Survived, gap_x = 0.04, gap_y = 0
  ) +
  scale_x_marimekko() +
  labs(title = "Wide column gaps, no vertical gaps")
```

```{r gap-xy2}
ggplot(titanic) +
  geom_marimekko(aes(fill = Survived, weight = Freq),
    formula = ~ Class | Survived, gap_x = 0, gap_y = 0.03
  ) +
  scale_x_marimekko() +
  labs(title = "No column gaps, visible vertical gaps")
```

## Plotly interactivity

marimekko plots work with `plotly::ggplotly()` out of the box:

```{r plotly, eval = FALSE}
library(plotly)

p <- ggplot(titanic) +
  geom_marimekko(aes(fill = Survived, weight = Freq),
    formula = ~ Class | Survived
  ) +
  scale_x_marimekko()
ggplotly(p)
```

## In-aesthetic expressions

Unlike some mosaic packages, `marimekko` supports arbitrary R
expressions — both in formulas and inside `aes()`:

```{r in-aes}
# Expressions work in formulas
ggplot(mtcars) +
  geom_marimekko(formula = ~ factor(cyl) | factor(gear)) +
  scale_x_marimekko() +
  labs(
    x = "Cylinders", fill = "Gears",
    title = "factor() inside formula works"
  )
```

## Namespace-qualified usage

`marimekko` works correctly when called with `::` notation
(e.g., `marimekko::geom_marimekko()`) without requiring
`library(marimekko)`. This makes it safe to use inside other packages
via `Imports` rather than `Depends`.

## Summary of parameters

| Parameter | Used in | Description |
|-----------|---------|-------------|
| `formula` | `geom_marimekko()` | Formula specifying variable hierarchy (`~ a \| b \| c`) |
| `gap` | `geom_marimekko()`, `geom_marimekko_text()`, `geom_marimekko_label()`, `geom_marimekko_jitter()`, `fortify_marimekko()` | Spacing between tiles (fraction of plot area) |
| `gap_x` | All geom/stat/fortify functions | Horizontal gap (overrides `gap` for x) |
| `gap_y` | All geom/stat/fortify functions | Vertical gap (overrides `gap` for y) |
| `standardize` | `geom_marimekko_text()`, `geom_marimekko_label()`, `geom_marimekko_jitter()`, `fortify_marimekko()` | Equal-width columns (spine plot) |
| `residuals` | `geom_marimekko_text()`, `geom_marimekko_label()`, `fortify_marimekko()` | Compute Pearson residuals |
| `seed` | `geom_marimekko_jitter()` | Reproducible jitter placement |
| `show_percentages` | `scale_x_marimekko()` | Append marginal % to x-axis labels |
